Polynomial Chaos for the Estimation of Manning's Friction

### 2013-11-13 paper/mcmc.tex
minor edits and added class files

Before attempting solving the inverse problem,
we need to make sure that the observed data are
comparable to the simulated one.
To this end, Figure~\ref{fig:compare}
compares PC mean water surface elevation profiles
with their counterparts from the observed gauge locations.
The plots show good agreement between the simulations and the observations at different
locations and at different times. 
%

Figure~\ref{fig:compare} can be recast as a scatter plot shown in 
Figure~\ref{fig:scatter}.

We now can exploit the surrogate model in the Bayesian inference of the Manning's 
roughness coefficient.  To this end, an adaptive MCMC method is used to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's roughness coefficient distributions in light of the 
observed data. This sampling, demanding tens of thousands of forward simulations, 
would have been prohibitively expensive in the
absence of the surrogate, as the generation of each sample would have required an
independent Geoclaw realization. The surrogate provides a computationally
efficient alternative, and requires only evaluating the PCE for different values 
of the seed $\xxi$. The setup of the Bayesian inference problem,
MCMC chains for the input drag parameters and the corresponding posterior distributions are 
presented and discussed in this section.

@@

The computed MCMC chains can be readily used to determine the posterior 
distribution; kernel density estimation (KDE) is used for this purpose
~\citep{Parzen1962,Silverman1986}(The first 5000 iterates, associated 
with the burn-in period, are discarded). The resulting posterior pdfs 
of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
in Figure~\ref{fig:pdfs}.  As expected from the chains shown in Figure
~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
with a Maximum A Posteriori (MAP) estimate of around 0.013; an extended tail 
towards the higher Manning's roughness coefficient values is also observed.
In contrast, the posterior pdfs of $N_1$ and $N_3$ appear to be fairly flat, 
and similar to the uniform prior. This is an indication that 
the observed data were not useful to refine our prior knowledge for $N_1$ and $N_3$.

The posterior distribution of the variance is also shown in Figure~\ref{fig:mcmc}. 
Taking the square root of the MAP values yields the water surface elevation standard 
deviation, and the latter is a reflection of the mismatch between the model and 
observed data estimated to be 0.144~$m$.

   0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835

>>>>

Before attempting solving the inverse problem,
we need to make sure that the observed data are
comparable to the simulated one.
To this end, Figure~\ref{fig:compare}
compares PC mean water surface elevation profiles
with their counterparts from the observed gauge locations.
The plots show good agreement between the simulations and the observations at different
locations and at different times. Figure~\ref{fig:compare} can be recast as a scatter plot shown in 
Figure~\ref{fig:scatter}.

We now can exploit the surrogate model in the Bayesian inference of the Manning's 
roughness coefficient.  To this end, an adaptive MCMC method is used to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's roughness coefficient distributions in light of the 
observed data. This sampling, demanding tens of thousands of forward simulations, 
would have been prohibitively expensive in the
absence of the surrogate, as the generation of each sample would have required an
independent \geoclaw realization. The surrogate provides a computationally
efficient alternative, and requires only evaluating the PCE for different values 
of the seed $\xxi$. The setup of the Bayesian inference problem,
MCMC chains for the input drag parameters and the corresponding posterior distributions are 
presented and discussed in this section.

@@

The computed MCMC chains can be readily used to determine the posterior 
distribution; kernel density estimation (KDE) is used for this purpose
~\citep{Parzen1962,Silverman1986}(The first 5000 iterates, associated 
with the burn-in period, are discarded). The resulting posterior pdfs 
of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
in Figure~\ref{fig:pdfs}.  As expected from the chains shown in Figure
~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
with a Maximum A Posteriori (MAP) estimate of around 0.011; an extended tail 
towards the higher Manning's roughness coefficient values is also observed.
In contrast, the posterior pdfs of $N_1$ appear to be fairly flat, 
and similar to the uniform prior. This is an indication that 
the observed data were not useful to refine our prior knowledge for $N_1$.
While for $N_3$, we observed a posterior that has a well defined peak
of 0.185 but no clear pdf shape.
The posterior distribution of the variance is also shown in Figure~\ref{fig:mcmc}. 
Taking the square root of the MAP values yields the water surface elevation standard 
deviation, and the latter is a reflection of the mismatch between the model and 
observed data estimated to be 0.145~$m$.

   %MAPS 0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835

### 2013-11-13 paper/pcstudy.tex
minor edits and added class files

In order to check the consistency of the approximation, we compare
 water surface elevation from the realizations 
with those obtained from the PC surrogate. The different curves
show an excellent agreement (not shown) for all times. We also define
an error metric that measures the relative normalized root mean-square error between the left hand side function 
in Equation~(\ref{eq:stochseries})
and its PC representation at the sampling points:
\begin{equation} 
   E = \frac{\displaystyle
         \left(\sum_{\xxi \in \NISP} \left|U(\xxi) - \sum_{k = 0}^{P}
U_k\Psi_k(\xxi)\right|^2
         \right)^{1/2}}
        {\displaystyle
          \left(\sum_{\xxi \in \NISP} \left|U(\xxi)\right|^2\right)^{1/2} 
          },
\label{eq:error}
\end{equation}
where $\NISP$ is the 125-member ensemble obtained using the PC algorithm. 
This error metric calculated at the different gauge locations is shown in Figure~\ref{fig:error};
the largest relative normalized error for 
water surface elevation is about 0.1\% .

@@

A final check consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC representation.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3}

The different curves
correspond to increased order of PC, from 1 to 5; 
The plots indicate that the double peaked distributions are
well-resolved with order  4 but becomes weakly insenstive to further refinement 
as shown with order 5.

The various error metrics presented above 
provide confidence that the PC expansion is a faithful 
model surrogate.

\clearpage

>>>>

In order to check the consistency of the approximation, we compare
 water surface elevation from the realizations 
with those obtained from the PC surrogate. The different curves
show an excellent agreement (not shown) for all times. We also define
an error metric that measures the relative normalized root mean-square error between the left hand side function 
in Equation~(\ref{eq:stochseries})
and its PC representation at the sampling points:
\begin{equation} 
   E = \frac{\displaystyle
         \left(\sum_{\xxi \in \NISP} \left|U(\xxi) - \sum_{k = 0}^{P}
U_k\Psi_k(\xxi)\right|^2
         \right)^{1/2}}
        {\displaystyle
          \left(\sum_{\xxi \in \NISP} \left|U(\xxi)\right|^2\right)^{1/2} 
          },
\label{eq:error}
\end{equation}
where $\NISP$ is the 125-member ensemble obtained using the PC algorithm. 
This error metric calculated at the different gauge locations is shown in Figure~\ref{fig:error};
the largest relative normalized error for 
water surface elevation is about 0.1\%.

@@

A final check consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC representation.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3}
The different curves
correspond to increased order of PC, from 1 to 5.
The plots indicate that the double peaked distributions are
well-resolved with order  4 but becomes weakly insenstive to further refinement 
as shown with order 5. The various error metrics presented above 
provide confidence that the PC expansion is a faithful 
model surrogate.
